From Reversible Programs to \\ Univalent Universes and Back

### 2017-07-09 upi.tex
remove lemmas
Some comments in sec 3

@@

Propositional equality, or identity types in intensional dependent
type theory, or path types under the HoTT-interpretation, come equipped
with a single constructor \AgdaInductiveConstructor{refl} and an
induction principle $\AgdaFunction{J}$. A peculiar aspect of
$\AgdaFunction{J}$ is that it does \emph{not} enforce that every
element of an identity type
$\AgdaFunction{x}~\AgdaDatatype{≡}~\AgdaFunction{y}$ is
\AgdaInductiveConstructor{refl}. This ``weakness'' has a beautiful
consequence: instead of identifying all proofs and collapsing the
structure of proofs in the process, it instead permits a rich
combinatorial structure to emerge.

The second notion we need is that of equivalence. We say that two types or
spaces $A$ and $B$ are equivalent if there exist functions $f : A \rightarrow B$
and $g : B \rightarrow A$ that compose to the identity function.
\footnote{Strictly speaking, this is a \emph{quasi-equivalence}, an equivalence
  also requires a certain coherence between the compositions.}  A notable point
is that equivalences do not come with an induction principle.  We say a universe
$\mathcal{U}$ is univalent if its space of identities
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of equivalences
$\AgdaFunction{A} ≃ \AgdaFunction{B}$.

@@

@@

\[\def\arraystretch{1.2}\begin{array}{rcll}
\AgdaFunction{$\dbracket{ \_ }_1$}  &:& (\bt \leftrightarrow \bt) \rightarrow `\bt \equiv `\bt \\
\AgdaFunction{$\dbracket{ \id }_1$} &=& `\mathbf{id}\\
\AgdaFunction{$\dbracket{ \swap }_1$} &=& `\mathbf{not}\\
\AgdaFunction{$\dbracket{ ! p }_1$} &=& (\dbracket{ p }_1)^{-1} \\
\AgdaFunction{$\dbracket{ p \odot q }_1$} &=& \dbracket{ p }_1 \circ \dbracket{ p }_1\\
\end{array}
\]

\begin{lemma}
  For all 1-combinators $p$, $p \isotwo \dbracket{ \dbracket{ p }_1 }_1^{-1}$.
\end{lemma}

\begin{lemma}
  For all 2-combinators $\alpha : p \isotwo q$, $\alpha \isothree \dbracket{ \dbracket{ p }_1 }_1^{-1} \bullet \dbracket{ \dbracket{ \alpha }_2 }_2^{-1} \bullet (!! \; \dbracket{ \dbracket{ q }_1 }_1^{-1})$.
\end{lemma}

\begin{lemma}
  For all 1-path $p$, $p \equiv \dbracket{ \dbracket{ p }_1^{-1} }_1$.
\end{lemma}

\begin{lemma}
  For all 2-path $\alpha : p \equiv q$, $\alpha \equiv \dbracket{ \dbracket{ p }_1^{-1} }_1 \circ \dbracket{ \dbracket{ \alpha }_2^{-1} }_2 \circ (\dbracket{ \dbracket{ q }_1^{-1} }_1)^{-1}$.
\end{lemma}

>>>>

\newcommand{\newnote}[2]{{\sf {#1} $\clubsuit$ {#2} $\clubsuit$}}
\newcommand{\VC}[1]{{\color{orange}{\newnote{VC}{#1}}}}

\newcommand{\newtext}[1]{{\color{purple}{#1}}}

@@

Propositional equality \newtext{is embodied by} identity types in
intensional dependent type theory, or path types under the
HoTT-interpretation. The identity type comes equipped with a single
constructor \AgdaInductiveConstructor{refl} and an induction principle
$\AgdaFunction{J}$. A peculiar aspect of $\AgdaFunction{J}$ is that it
does \emph{not} enforce that every element of an identity type
$\AgdaFunction{x}~\AgdaDatatype{≡}~\AgdaFunction{y}$ is
\AgdaInductiveConstructor{refl}. \VC{why is that peculiar?}
\newtext{However, unlike other types, it is not necessary to have a
  uniqueness principle ($K$) for it, which facilitates the homotopy
  interpretation of intensional type theory.} This ``weakness'' has a
beautiful consequence: instead of identifying all proofs and collapsing
the structure of proofs in the process, it instead permits a rich
combinatorial structure to emerge. \VC{cite Homotopy Theoretic Aspects
  of Constructive Type Theory by Michael Warren}

The second notion we need is that of equivalence. We say that two types
or spaces $A$ and $B$ are \newtext{(homotopy)} equivalent if there exist
functions $f : A \rightarrow B$ and $g : B \rightarrow A$ that compose
to the identity function. \footnote{Strictly speaking, this is a
  \emph{quasi-equivalence}, an equivalence also requires a certain
  coherence between the compositions.}  A notable point is that
equivalences do not come with an induction principle. \VC{What does this
  mean exactly? Maybe you're trying to say that it does not satisfy a
  principle like path induction, but in a univalent universe, it will!}
We say a universe $\mathcal{U}$ is univalent if its space of identities
$\AgdaFunction{A} ≡ \AgdaFunction{B}$ is equivalent to its space of
equivalences $\AgdaFunction{A} ≃ \AgdaFunction{B}$.

@@

\VC{Explain universal fibrations, object classifier?}

@@
